\bf Model Visualization Techniques for a Social Network Model

### 2017-04-03 paper.Rnw
add seriation ex.

A key component of each step of the SIENA method of moments algorithm is the ``microstep." This step simulates from the model in its current state, $x(t_m)$ with current parameter values $\theta_{0}$, to the next state, $x(t_{m+1})$. This microstep process stops when the simulated network has acheived the same number of differences, $C$, from $x(t_m)$ as $x(t_{m+1})$, where 
$$C = \sum_{i \neq j} |x_{ij}(t_{m+1}) - x_{ij}(t_{m})|.$$ 
This simulation process follows the steps of the continuous-time Markov chain. First, an ``ego node'' is selected to make a change, and then it randomly makes one change in its ties according to the probability, $p_{ij}$, determined by its objective function. The options for change are (1) removing a current tie, (2) adding a new tie, or (3) making no change at all. Saving all of these steps is not efficient if you're only interested in parameter values, but they can be saved and extracted using options in \texttt{RSiena}, which is exactly what we did to create our visualizations. Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps. We wanted to see these in-between steps in order to better understand the behavior of the underlying continuous-time Markov chain.

@@

@@

INSERT DESCRIPTION OF THE MOVIE - WHAT IT ALLOWS US TO SEE, ETC.

@@

To reorder the vertices for the matrix visualization, we first constructed a cumulative adjaceny matrix, $\mathbf{A}^{cum}$, for the series of microsteps simulating the network from $x(t_m)$ to $x(t_{m+1})$. A single entry in the cumulative adjacency matrix, $\mathbf{A}^{cum}_{ij}$, is the total number of times the edge from node $i$ to node $j$ appears in the network from the initial observation, $x(t_m) \equiv X(0)$ to the final reult of the last microstep, $X(C)$: 
$$ \mathbf{A}^{cum}_{ij} = \sum_{c = 0}^C X_{ij}(c). $$
We then performed a principal component analysis on $\mathbf{A}^{cum}$, and used the values of the first principal component to order the vertices on the $x$ and $y$ axes for the adjacency matrix animation.

>>>>

A key component of each step of the SIENA method of moments algorithm is the ``microstep." This step simulates from the model in its current state, $x(t_m)$ with current parameter values $\theta_{0}$, to the next state, $x(t_{m+1})$. This microstep process stops when the simulated network has acheived the same number of differences, $C$, from $x(t_m)$ as $x(t_{m+1})$, where 
$$C = \sum_{i \neq j} |x_{ij}(t_{m+1}) - x_{ij}(t_{m})|.$$ 
This simulation process follows the steps of the continuous-time Markov chain. First, an ``ego node'' is selected to make a change, and then it randomly makes one change in its ties according to the probability, $p_{ij}$, determined by its objective function. The options for change are (1) removing a current tie, (2) adding a new tie, or (3) making no change at all. Saving all of these steps is not efficient if you're only interested in estimating parameter values, but they can be saved and extracted using options in \texttt{RSiena}, which is exactly what we did to create our visualizations. Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps, depending on the size of the network. We wanted to see these in-between steps in order to better understand the behavior of the underlying continuous-time Markov chain.

@@

<<makinggif>>=
library(RSiena)
library(tidyverse)
library(sna)
library(network)
library(netvizinf)
library(gganimate)
friend.data.w1 <- as.matrix(read.table("../data/s50-network1.dat"))
friend.data.w2 <- as.matrix(read.table("../data/s50-network2.dat"))
friend.data.w3 <- as.matrix(read.table("../data/s50-network3.dat"))
drink <- as.matrix(read.table("../data/s50-alcohol.dat"))
fd2.w1 <- friend.data.w1[20:35,20:35]
fd2.w2 <- friend.data.w2[20:35,20:35]
fd2.w3 <- friend.data.w3[20:35,20:35]
friendshipData <- array(c(fd2.w1, fd2.w2,fd2.w3), dim = c(16, 16, 3))

# running Siena
friendship <- sienaDependent(friendshipData)
alcohol <- varCovar(drink[20:35,])
mydata <- sienaDataCreate(friendship, alcohol)
myeffnull <- getEffects(mydata)
myalgorithm <- sienaAlgorithmCreate(projname = 's16_3')
set.seed(823746)
ansnull <- siena07(myalgorithm, data = mydata, effects = myeffnull,
                   returnChains = TRUE, returnDataFrame = TRUE,
                   returnDeps = TRUE, silent = TRUE, verbose = FALSE,
                   batch = TRUE)
ansnullchains <- get_chain_info(ansnull)
ansnullchains %>%
  filter(period == 1) %>%  #only look at chains from wave 1 to wave 2
  group_by(rep) %>%
  select(rep, from = ego, to = alter) %>%
  mutate(val = as.numeric(!from == to),
         from = paste0("V", parse_number(from)+1), # make the chains
         to = paste0("V", parse_number(to)+1)) -> ansnullchainsw1w2
library(geomnet)
colnames(fd2.w1) <- paste0("V", 1:16)
rownames(fd2.w1) <- paste0("V", 1:16)
wave1friends <- fortify(as.adjmat(fd2.w1))
ms1 <- listMicrosteps(dat = wave1friends,
                      microsteps = filter(ansnullchainsw1w2, rep == 1))

@@

XX INSERT DESCRIPTION OF THE MOVIE - WHAT IT ALLOWS US TO SEE, ETC. XX

@@

To reorder the vertices for the matrix visualization, we first constructed a cumulative adjaceny matrix, $\mathbf{A}^{cum}$, for the series of microsteps simulating the network from $x(t_m)$ to $x(t_{m+1})$. A single entry in the cumulative adjacency matrix, $\mathbf{A}^{cum}_{ij}$, is the total number of times the edge from node $i$ to node $j$ appears in the network from the initial observation, $x(t_m) \equiv X(0)$ to the final reult of the last microstep, $X(C)$: 
$$ \mathbf{A}^{cum}_{ij} = \sum_{c = 0}^C X_{ij}(c). $$
We then performed a principal component analysis on $\mathbf{A}^{cum}$, and used the values of the first principal component to order the vertices on the $x$ and $y$ axes for the adjacency matrix animation. For one such series of microsteps simulated by \texttt{RSiena}, we present the adjacency matrix ordered by the (arbitrary) vertex id alongside the seriated adjacency microsteps using the first principal component loading on the cumulative adjacency matrix, $A^{cum}$

<<adjmatorder, fig.cap="On the left, the starting friendship network represented in adjacency matrix form, ordered by vertex id. On the right, the same adjacency matrix is presented after ordering the vertices by one repitition of the microstep simulation process from wave one to wave two.", fig.width=8>>=
# first present the original wave 1 adj mat, then present the adj mat after being ordered by the 1st pc of the big adjmat A. 
numordersf <- data.frame(fd2.w1) %>% 
  mutate(from = rownames(fd2.w1)) %>% 
  gather(key = to ,value = value, -from)
pcaordersf <- numordersf

numordersf$from <- factor(numordersf$from, levels = paste0("V", 1:16))
numordersf$to <- factor(numordersf$to, levels = paste0("V", 1:16))

p1 <- ggplot(data = numordersf, aes(x = from, y = to)) + geom_tile(color = 'black', aes(fill = as.factor(value))) + 
  scale_fill_manual(values = c('white', 'black')) +
  theme(legend.position = 'none', axis.text = element_text(size = 8)) + 
  coord_fixed()

sfmsall <- read_csv("data/smallfriendsmicrosteps.csv")

sfmsall %>% 
  group_by(from, to) %>% 
  filter(rep == 1) %>% 
  summarise(count = n()) -> sfmsall_summ
sfmsall_summ$from <- factor(sfmsall_summ$from, levels = paste0('V', 1:16))
sfmsall_summ$to <- factor(sfmsall_summ$to, levels = paste0('V', 1:16))
sfmsall_summ %>% 
  spread(to, count, fill = 0) %>% 
  data.frame -> allmsrep1_adjmat
allmsrep1_adjmat$V16 <- 0
allmsrep1_adjmat <- allmsrep1_adjmat[,-1]
allmsrep1_adjmat <- as.matrix(allmsrep1_adjmat)
rownames(allmsrep1_adjmat) <- paste0("V", 1:16)

pca1 <- prcomp(allmsrep1_adjmat)
order1 <- names(sort(pca1$rotation[,1], decreasing = T))
order2 <- names(sort(abs(pca1$rotation[,1]), decreasing = T))

pcaordersf$from <- factor(pcaordersf$from, levels = order1)
pcaordersf$to <- factor(pcaordersf$to, levels = order1)

p2 <- ggplot(data = pcaordersf, aes(x = from, y = to)) + geom_tile(color = 'black', aes(fill = as.factor(value))) + 
  scale_fill_manual(values = c('white', 'black')) +
  theme(legend.position = 'none', axis.text = element_text(size = 8)) + 
  coord_fixed()

grid.arrange(p1,p2, nrow = 1)
@
